---
title: "Blog"
output: html_document
date: "2025-09-09"
---

# 1. January

# Calendar

This post is to create a calendar to check when some events were more likely to occur.

# Packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Data

Create a data frame with a sequence of days

```{r}
firstday<- as.Date("2020-05-01")
lastday <- as.Date("2020-08-31")
days_seq <- seq(firstday, lastday, by = "day")
days_df<-as.data.frame(days_seq)
```

Separate months and days

```{r}
days_df$month<-substr(days_df$days, start = 6, stop = 7)
days_df$day<-substr(days_df$days, start = 9, stop = 10)
```

Add values

```{r}
my_values<-sample(1:7, 123, replace=TRUE)
days_df$events<-as.factor(my_values[1:123])
```

Separate months

```{r}
may<-days_df %>%
  filter(month=='05')
```

Create x and y axis

```{r}
calen_xs<-rep(c(1:7), times = 5)
calen_ys<-rep(c(5:1), each = 7)
```

```{r}
may$calen_xs<-calen_xs[1:31]
may$calen_ys<-calen_ys[1:31]
```

Select color palette

```{r}
my_palette <- c("1" = '#577590', 
                "2" = '#43aa8b', 
                "3" = '#90be6d', 
                "4" = "#f9c74f", 
                "5" = "#f8961e", 
                "6" = "#f3722c", 
                "7" = "#f94144") 
```

# Plot

```{r, fig.height=4.8, fig.width=6.1}
ggplot(may,aes(x=calen_xs,y=calen_ys,color=events))+
  geom_point(size=25,shape=15)+

  scale_x_continuous(limits=c(0.5,7.5))+
  scale_y_continuous(limits=c(0.5,5.5))+
  
  # remove background colors
  theme_void()+
  
  # adds texts
  geom_text(aes(label=day),color='white') +
  
  # sets the legend below
  theme(legend.position = 'top')+
   
  # lines the legend
  guides(colour = guide_legend(override.aes = list(size=6),nrow = 1))+
  
  # uses palette
  scale_color_manual(name = "May no. events",values = my_palette)
```

# Thats it!

# Export

```{r, echo=FALSE, eval=FALSE}
library(here)
here()
ggsave(ggplot(may,aes(x=calen_xs,y=calen_ys,color=events))+
  geom_point(size=25,shape=15)+

  scale_x_continuous(limits=c(0.5,7.5))+
  scale_y_continuous(limits=c(0.5,5.5))+
  
  # remove background colors
  theme_void()+
  
  # adds texts
  geom_text(aes(label=day),color='white') +
  
  # sets the legend below
  theme(legend.position = 'top')+
   
  # lines the legend
  guides(colour = guide_legend(override.aes = list(size=6),nrow = 1))+
  
  # uses palette
  scale_color_manual(name = "May no. events",values = my_palette),
       filename = paste0(here(),"/Blog/2024/calendar_export.jpg"),
       width = 6, 
       height = 7, 
       units = "in",
       dpi = 300)
```

# ~~~~~~~~~~~~~
# 2. February
# Environmental variables

# Intro

This post is to create a map with environmental variables.

1.  Identify the data you want to download
2.  Download data from the server
3.  Create a basic map with the values

# Define

Define the area you want to download the data from.

```{r}
xlim1<--125.
xlim2<--88.
ylim1<-14.
ylim2<-35.
```

Define also the time frame you are interested in.

```{r}
time1<-'2017-03-01T00:00:00Z'
time2<-'2017-03-30T00:00:00Z'
```

# Load data

Load the package *rerddap*

```{r}
library("rerddap")
```

Check the list of available servers.

For the example, we are interested in the Pacific Ocean, therefore I would use NOAA.

```{r, eval=FALSE}
servers()
```

I want to download sea surface temperature (SST) data, therefore I will use *erdMH1sstdmday*

```{r}
sstInfo <- info('erdMH1sstdmday')
```

To check the specifications use *browse*

```{r, eval=FALSE}
browse('erdMH1sstdmday') 
```

It takes some time to download.

```{r}
SST03.2017<-griddap(sstInfo, latitude = c(ylim1, ylim2), longitude = c(xlim1, xlim2), time = c(time1,time2), fields = 'sst')
```

To use ggplot, the format in dataframe works better.

```{r}
SST03.2017dt<-SST03.2017$data
```

Clear NaNs

```{r, warning=FALSE}
library(tidyverse)
```

```{r}
SST03.2017dt_clean<-SST03.2017dt %>%
  filter(sst!='NaN')
```

Create plot

```{r}
ggplot(SST03.2017dt_clean) +
 geom_raster(aes(x=longitude, y=latitude, fill = sst))+ 
  scale_fill_viridis_c(option = "H")
```

# Add land for reference

Load the package *rworldmap* this includes shapefiles of countries

```{r}
library(rworldmap)
```

The function *getMap()* loads the map in your environment

```{r}
worldMap <- getMap()
```

Load the package tidyverse

Use the function *fortify* to be able to plot the map using ggplot

```{r}
world.points <- fortify(worldMap)
```

To plot using ggplot a data frame is recommended

```{r}
world.points$region <- world.points$id
world.df <- world.points[,c("long","lat","group", "region")]
```

Create plot using ggplot

```{r}
ggplot(SST03.2017dt) +
  geom_raster(aes(x=longitude, y=latitude, fill = sst))+
  geom_polygon(data = world.df, aes(x = long, y = lat, group = group), colour = '#403d39', fill = "#e5e5e5") +
  theme(
    axis.title = element_blank(),
    panel.border = element_rect(colour = "#495057", fill=NA, size=1),
    panel.grid = element_blank(), 
    panel.spacing = unit(0, "lines"), 
    plot.background = element_blank())+
    coord_sf(xlim = c(xlim1+2,xlim2-2), 
             ylim = c(ylim1+2,ylim2-2))+ 
  scale_fill_viridis_c(option = "H")
```

```{r, echo=FALSE, eval=FALSE}
library(here)
here()
ggsave(ggplot(SST03.2017dt) +
  geom_raster(aes(x=longitude, y=latitude, fill = sst))+
  geom_polygon(data = world.df, aes(x = long, y = lat, group = group), colour = '#403d39', fill = "#e5e5e5") +
  theme(
    axis.title = element_blank(),
    panel.border = element_rect(colour = "#495057", fill=NA, size=1),
    panel.grid = element_blank(), 
    panel.spacing = unit(0, "lines"), 
    plot.background = element_blank())+
    coord_sf(xlim = c(xlim1+2,xlim2-2), 
             ylim = c(ylim1+2,ylim2-2))+ 
  scale_fill_viridis_c(option = "H"), 
       filename = paste0(here::here(),"/Blog/2024/environmental_export.jpg"),
       width = 8, 
       height = 8, 
       units = "in",
       dpi = 300)
```


# ~~~~~~~~~~~~~
# 3. March
# Double axis

# Intro

This post is to create a plot with environmental variables, in one axis the sea surface temperature and in the other axis the chlorophyll-a concentration.

1.  Download data from the server
2.  Calculate average values
3.  Create a secondary-axis plot with the values

# Load data

Load the package **rerddap**

```{r}
library("rerddap")
```

**SST: sea surface temperature**

For obtaining data, download sea surface temperature (SST) from **erdMWsstd1day_LonPM180** ([link here](https://upwell.pfeg.noaa.gov/erddap/griddap/erdMWsstd1day_LonPM180.graph?sst%5B(2024-03-10T12:00:00Z)%5D%5B(0.0)%5D%5B(23):(27)%5D%5B(-111):(-109)%5D&.draw=surface&.vars=longitude%7Clatitude%7Csst&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff)).

Another option is **erdMW1sstd1day** ([link here](https://upwell.pfeg.noaa.gov/erddap/griddap/erdMWsstd1day.graph?sst%5B(2024-02-17T12:00:00Z)%5D%5B(0.0)%5D%5B(22.0):(51.0)%5D%5B(205.0):(255.0)%5D&.draw=surface&.vars=longitude%7Clatitude%7Csst&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff&.click=?302,181)).

To search other options [here](https://upwell.pfeg.noaa.gov/erddap/index.html).

```{r}
sstInfo <- info('erdMWsstd1day_LonPM180')
```

To subset the data select the coordinates of a smaller area.

```{r}
lonmin<--111
lonmax<--109
latmin<-23
latmax<-27
```

To download, provide the parameters such as the server, the coordinates and the time frame. It takes some time to download.

```{r}
SST03.2023<-griddap(sstInfo, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-03-01T00:00:00Z','2023-03-30T00:00:00Z'), 
                    fields = 'sst')

SST04.2023<-griddap(sstInfo, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-04-01T00:00:00Z','2023-04-30T00:00:00Z'), 
                    fields = 'sst')

SST05.2023<-griddap(sstInfo, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-05-01T00:00:00Z','2023-05-30T00:00:00Z'),
                    fields = 'sst')

SST06.2023<-griddap(sstInfo, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-06-01T00:00:00Z','2023-06-30T00:00:00Z'), 
                    fields = 'sst')

SST07.2023<-griddap(sstInfo, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-07-01T00:00:00Z','2023-07-30T00:00:00Z'), 
                    fields = 'sst')

SST08.2023<-griddap(sstInfo, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-08-01T00:00:00Z','2023-08-30T00:00:00Z'), 
                    fields = 'sst')

SST09.2023<-griddap(sstInfo, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-09-01T00:00:00Z','2023-09-30T00:00:00Z'), 
                    fields = 'sst')

SST10.2023<-griddap(sstInfo, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-10-01T00:00:00Z','2023-10-30T00:00:00Z'), 
                    fields = 'sst')

SST11.2023<-griddap(sstInfo, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-11-01T00:00:00Z','2023-11-30T00:00:00Z'), 
                    fields = 'sst')

SST12.2023<-griddap(sstInfo, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-12-01T00:00:00Z','2023-12-30T00:00:00Z'), 
                    fields = 'sst')
```

**CHL: chlorophyll-a concentration**

To download chlorophyll concentration (CHL) data, connect to **erdMWchlamday_LonPM180** (here the [link](https://upwell.pfeg.noaa.gov/erddap/griddap/erdMWchlamday_LonPM180.html).

Another option is **erdMW1CHLd1day** (here the [link](https://upwell.pfeg.noaa.gov/erddap/griddap/erdMWcflh3day.graph))

```{r}
CHLInfo <- info('erdMWchlamday_LonPM180')
```

To subset the data select the coordinates of a smaller area.

To download, provide the parameters such as the server, the coordinates and the time frame. It takes some time to download.

```{r}
CHL03.2023<-griddap(CHLInfo, 
                    latitude = c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-03-01T00:00:00Z','2023-03-30T00:00:00Z'), 
                    fields = 'chlorophyll')

CHL04.2023<-griddap(CHLInfo, 
                    latitude = c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-04-01T00:00:00Z','2023-04-30T00:00:00Z'), 
                    fields = 'chlorophyll')

CHL05.2023<-griddap(CHLInfo, 
                    latitude = c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-05-01T00:00:00Z','2023-05-30T00:00:00Z'), 
                    fields = 'chlorophyll')
CHL06.2023<-griddap(CHLInfo, 
                    latitude = c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-06-01T00:00:00Z','2023-06-30T00:00:00Z'), 
                    fields = 'chlorophyll')

CHL07.2023<-griddap(CHLInfo, 
                    latitude = c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-07-01T00:00:00Z','2023-07-30T00:00:00Z'), 
                    fields = 'chlorophyll')

CHL08.2023<-griddap(CHLInfo, 
                    latitude = c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-08-01T00:00:00Z','2023-08-30T00:00:00Z'), 
                    fields = 'chlorophyll')

CHL09.2023<-griddap(CHLInfo, 
                    latitude = c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-09-01T00:00:00Z','2023-09-30T00:00:00Z'), 
                    fields ='chlorophyll')

CHL10.2023<-griddap(CHLInfo, 
                    latitude = c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-10-01T00:00:00Z','2023-10-30T00:00:00Z'), 
                    fields ='chlorophyll')

CHL11.2023<-griddap(CHLInfo, 
                    latitude = c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-11-01T00:00:00Z','2023-11-30T00:00:00Z'), 
                    fields = 'chlorophyll')

CHL12.2023<-griddap(CHLInfo, 
                    latitude = c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-12-01T00:00:00Z','2023-12-30T00:00:00Z'), 
                    fields = 'chlorophyll')
```

# Average values

Load the package **tidyverse**.

```{r}
library(tidyverse)
```

**SST: sea surface temperature**

To calculate average values the best is to transform the data into a data frame.

```{r}
SST03.2023dt<-SST03.2023$data
SST04.2023dt<-SST04.2023$data
SST05.2023dt<-SST05.2023$data
SST06.2023dt<-SST06.2023$data
SST07.2023dt<-SST07.2023$data
SST08.2023dt<-SST08.2023$data
SST09.2023dt<-SST09.2023$data
SST10.2023dt<-SST10.2023$data
SST11.2023dt<-SST11.2023$data
SST12.2023dt<-SST12.2023$data
```

Functions from the package tidyverse can be used to calculate the **mean** values.\
The xaxis would be in numeric value for the plot.

```{r}
SST03.2023dt_clean<-SST03.2023dt %>%  drop_na(sst) %>% 
  summarise(mean_sst=mean(sst),sd_sst=sd(sst))%>% 
  mutate(month='03/2023', xaxis=1)

SST04.2023dt_clean<-SST04.2023dt %>%  drop_na(sst) %>% 
  summarise(mean_sst=mean(sst),sd_sst=sd(sst))%>% 
  mutate(month='04/2023', xaxis=2)

SST05.2023dt_clean<-SST05.2023dt %>%  drop_na(sst) %>% 
  summarise(mean_sst=mean(sst),sd_sst=sd(sst))%>% 
  mutate(month='05/2023', xaxis=3)

SST06.2023dt_clean<-SST06.2023dt %>%  drop_na(sst) %>% 
  summarise(mean_sst=mean(sst),sd_sst=sd(sst))%>% 
  mutate(month='06/2023', xaxis=4)

SST07.2023dt_clean<-SST07.2023dt %>%  drop_na(sst) %>% 
  summarise(mean_sst=mean(sst),sd_sst=sd(sst))%>% 
  mutate(month='07/2023', xaxis=5)

SST08.2023dt_clean<-SST08.2023dt %>%  drop_na(sst) %>% 
  summarise(mean_sst=mean(sst),sd_sst=sd(sst))%>% 
  mutate(month='08/2023', xaxis=6)

SST09.2023dt_clean<-SST09.2023dt %>%  drop_na(sst) %>% 
  summarise(mean_sst=mean(sst),sd_sst=sd(sst))%>% 
  mutate(month='09/2023', xaxis=7)

SST10.2023dt_clean<-SST10.2023dt %>%  drop_na(sst) %>% 
  summarise(mean_sst=mean(sst),sd_sst=sd(sst))%>% 
  mutate(month='10/2023', xaxis=8)

SST11.2023dt_clean<-SST11.2023dt %>%  drop_na(sst) %>% 
  summarise(mean_sst=mean(sst),sd_sst=sd(sst))%>% 
  mutate(month='11/2023', xaxis=9)

SST12.2023dt_clean<-SST12.2023dt %>%  drop_na(sst) %>% 
  summarise(mean_sst=mean(sst),sd_sst=sd(sst))%>% 
  mutate(month='12/2023', xaxis=10)
```

Now join into a single data frame.

```{r}
SST_months<-rbind(SST03.2023dt_clean,
      SST04.2023dt_clean,
      SST05.2023dt_clean,
      SST06.2023dt_clean,
      SST07.2023dt_clean,
      SST08.2023dt_clean,
      SST09.2023dt_clean,
      SST10.2023dt_clean,
      SST11.2023dt_clean,
      SST12.2023dt_clean
      )
```

**CHL: chlorophyll-a concentration**

The same procedure used with SST, would be used with CHL.

Convert to data frame.

```{r}
CHL03.2023dt<-CHL03.2023$data
CHL04.2023dt<-CHL04.2023$data
CHL05.2023dt<-CHL05.2023$data
CHL06.2023dt<-CHL06.2023$data
CHL07.2023dt<-CHL07.2023$data
CHL08.2023dt<-CHL08.2023$data
CHL09.2023dt<-CHL09.2023$data
CHL10.2023dt<-CHL10.2023$data
CHL11.2023dt<-CHL11.2023$data
CHL12.2023dt<-CHL12.2023$data
```

Calculate the mean values. Include the xaxis.

```{r}
CHL03.2023dt_clean<-CHL03.2023dt %>%  drop_na(chlorophyll) %>% 
  summarise(mean_CHL=mean(chlorophyll),sd_CHL=sd(chlorophyll))%>%  
  mutate(month='03/2023', xaxis=1)

CHL04.2023dt_clean<-CHL04.2023dt %>%  drop_na(chlorophyll) %>%
  summarise(mean_CHL=mean(chlorophyll),sd_CHL=sd(chlorophyll))%>%
  mutate(month='04/2023', xaxis=2)

CHL05.2023dt_clean<-CHL05.2023dt %>%  drop_na(chlorophyll) %>% 
  summarise(mean_CHL=mean(chlorophyll),sd_CHL=sd(chlorophyll))%>%
  mutate(month='05/2023', xaxis=3)

CHL06.2023dt_clean<-CHL06.2023dt %>%  drop_na(chlorophyll) %>% 
  summarise(mean_CHL=mean(chlorophyll),sd_CHL=sd(chlorophyll))%>%
  mutate(month='06/2023', xaxis=4)

CHL07.2023dt_clean<-CHL07.2023dt %>%  drop_na(chlorophyll) %>% 
  summarise(mean_CHL=mean(chlorophyll),sd_CHL=sd(chlorophyll))%>%
  mutate(month='07/2023', xaxis=5)

CHL08.2023dt_clean<-CHL08.2023dt %>%  drop_na(chlorophyll) %>% 
  summarise(mean_CHL=mean(chlorophyll),sd_CHL=sd(chlorophyll))%>%
  mutate(month='08/2023', xaxis=6)

CHL09.2023dt_clean<-CHL09.2023dt %>%  drop_na(chlorophyll) %>% 
  summarise(mean_CHL=mean(chlorophyll),sd_CHL=sd(chlorophyll))%>%
  mutate(month='09/2023', xaxis=7)

CHL10.2023dt_clean<-CHL10.2023dt %>%  drop_na(chlorophyll) %>% 
  summarise(mean_CHL=mean(chlorophyll),sd_CHL=sd(chlorophyll))%>%
  mutate(month='10/2023', xaxis=8)
CHL11.2023dt_clean<-CHL11.2023dt %>%  drop_na(chlorophyll) %>% 
  summarise(mean_CHL=mean(chlorophyll),sd_CHL=sd(chlorophyll))%>%
  mutate(month='11/2023', xaxis=9)

CHL12.2023dt_clean<-CHL12.2023dt %>%  drop_na(chlorophyll) %>% 
  summarise(mean_CHL=mean(chlorophyll),sd_CHL=sd(chlorophyll))%>%
  mutate(month='12/2023', xaxis=10)
```

Create a data frame with all the values.

```{r}
CHL_months<-rbind(CHL03.2023dt_clean,
      CHL04.2023dt_clean,
      CHL05.2023dt_clean,
      CHL06.2023dt_clean,
      CHL07.2023dt_clean,
      CHL08.2023dt_clean,
      CHL09.2023dt_clean,
      CHL10.2023dt_clean,
      CHL11.2023dt_clean,
      CHL12.2023dt_clean
      )
```

# Two-axis plot

```{r, eval=FALSE, echo=FALSE}
ggplot(SST_months, aes(x=month, y=mean_sst)) + 
  geom_line() +
  geom_point()+
  geom_errorbar(aes(ymin=mean_sst-sd_sst, ymax=mean_sst+sd_sst), width=.2,
                 position=position_dodge(0.05))
ggplot(CHL_months, aes(x=month, y=mean_CHL)) + 
  geom_line() +
  geom_point()+
  geom_errorbar(aes(ymin=mean_CHL-sd_CHL, ymax=mean_CHL+sd_CHL), width=.2,
                 position=position_dodge(0.05))
```

**SST: sea surface temperature**

Create a plot using **geom_line** using functions from the package ggplot.

```{r, fig.height=4, fig.width=6}
plot_SST<-ggplot(SST_months) +
  geom_line(aes(x = xaxis, y = mean_sst),color='red', size=1)+ 
  scale_y_continuous(("Sea Surface Temperature (C)"),limits=c(8.5,35.5),expand=c(0, 0))+
    scale_x_continuous("Months",breaks=c(1,2,3,4,5,6,7,8,9,10),labels=c('Mar','Apr','May','Jun','Jul','Ago','Sep','Oct','Nov','Dec'))+
  theme_bw()+#dejar lineas de las orillas
  theme(axis.title.x = element_text(size=13),
        axis.text.x = element_text(size=10),
        axis.title.y=element_text(size=13),
        axis.text.y=element_text(size=10),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  NULL
plot_SST
```

**CHL: chlorophyll-a concentration**

Create your second plot with functions from the package ggplot.

For this plot, as it would be on top of the other, the panel background, border, and plot background should be transparent.

```{r, fig.height=4, fig.width=6}
plot_CHL<-ggplot(CHL_months) +
  geom_line(aes(x = xaxis, y = mean_CHL),color='blue', size=1) +
  scale_y_continuous(("Chlorophyll-a concentration (mg m-3 )"),limits=c(0.05,3.0),breaks=c(0.5,1,1.5,2,2.5),expand=c(0, 0),position = "right")+
  scale_x_continuous("Months",breaks=c(1,2,3,4,5,6,7,8,9,10),labels=c('Mar','Apr','May','Jun','Jul','Ago','Sep','Oct','Nov','Dec'))+
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y = element_text(size = 13, vjust=6), 
        axis.text.y = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill='transparent'),
        panel.border = element_rect(colour = "black", fill='transparent', size=1),
        plot.background = element_rect(fill = 'transparent',colour = 'transparent',linewidth = 1))
plot_CHL
```

# Add legend

Using **annotate** the legend would be included inside the plot.

```{r, fig.height=4, fig.width=6}
plot_SST_wlegend<-plot_SST+
  annotate("text", x = c(1.9), y = c(34), label = c("SST") , color="black", size=5)+ 
  annotate("segment", x = 1.0, xend = 1.4, y = 34, yend = 34, colour = "red", size=1, alpha=1)+
  
  annotate("text", x = c(1.9), y = c(32), label = c("CHL") , color="black", size=5)+ 
  annotate("segment", x = 1.0, xend = 1.4, y = 32, yend = 32, colour = "blue", size=1, alpha=1)+
  NULL
plot_SST_wlegend
```

# Join graphs

Using functions from the package **patchwork** both plots would be merge into one.

```{r}
library(patchwork)
```

To further learn how to change the layout [click here](https://patchwork.data-imaginist.com/articles/guides/layout.html).

```{r, fig.height=4, fig.width=6}
plot_SST_wlegend + plot_CHL+
  plot_layout(design = c(area(t = 1, l = 1, b = 5, r = 4),
            area(t = 1, l = 1, b = 5, r = 4)))
```

```{r, echo=FALSE, eval=FALSE}
library(here)
here()
ggsave(plot_SST_wlegend + plot_CHL+
  plot_layout(design = c(area(t = 1, l = 1, b = 5, r = 4),
            area(t = 1, l = 1, b = 5, r = 4))), 
       filename = paste0(here::here(),"/Blog/2024/sst_chl_yaxis_export.jpg"),
       width = 8, 
       height = 8, 
       units = "in",
       dpi = 300)
```

# ~~~~~~~~~~~~~
# 4. April
# Buffer

In this blog, a buffer subsetting environmental data would be created and the average values within the buffer calculated.

# Download data

Load the package **rerddap**

```{r}
library("rerddap")
```

```{r}
SST_info <- info('erdMWsstd1day_LonPM180')
```

To subset the data select the coordinates of a smaller area.

```{r}
lonmin<--111
lonmax<--109
latmin<-23
latmax<-27
```

To download, provide the parameters such as the server, the coordinates and the time frame. It takes some time to download.

```{r}
SST_values<-griddap(SST_info, 
                    latitude= c(latmin, latmax), longitude = c(lonmin, lonmax), 
                    time = c('2023-03-01T00:00:00Z','2023-03-30T00:00:00Z'), 
                    fields = 'sst')
```

To access the data as data frame, extract the data.

```{r}
SST_df<-SST_values$data
```

Using functions from the package tidyverse, remove all NAs.

```{r, warning=FALSE}
library(tidyverse)
```

```{r}
SST_dfclean<-SST_df %>%
  filter(sst!='NaN')
```

# Map

Check data downloaded using ggplot.

```{r}
library(ggplot2)
```

Download data from land from the package **rworldmap**.

```{r}
library(rworldmap)
```

The function *getMap()* loads the map in your environment.

```{r}
world_map <- getMap()
```

To plot using ggplot a data frame is recommended. Use the function *fortify* to be able to get the data in a data frame format.

```{r, warning=FALSE}
world_points <- fortify(world_map)
world_points$region <- world_points$id
world_df <-world_points[,c("long","lat","group", "region")]
```

```{r, fig.height=4}
ggplot() +
 geom_raster(data=SST_dfclean,aes(x=longitude, y=latitude, fill = sst))+ scale_fill_viridis_c(option = "H")+
 geom_polygon(data = world_df, aes(x = long, y = lat, group = group), colour = '#403d39', fill = "#e5e5e5") +
 coord_sf(xlim = c(-109.7,-109.1),ylim = c(24.6,25.2))
```

# Create buffer

Load package **sp**.

```{r}
library(sp)
```

Give coordinates from the central point.

```{r}
This_point<-data.frame(Longitude=-109.37728617302638,Latitude=24.92572600240793)
```

Use this custom function to create a buffer. Note that it uses epsg 4236, adjust if necessary.

```{r}
create_buffer<-function(central_point=central_point, buffer_km=buffer_km){
  central_spatial<- sp::SpatialPoints(cbind(central_point$Longitude,central_point$Latitude)) 
  sp::proj4string(central_spatial)= sp::CRS("+init=epsg:4326") 
  central_spatial <- sp::spTransform(central_spatial, sp::CRS("+init=epsg:4326"))
  central_spatial<-sf::st_as_sf(central_spatial)
  buffer_dist<-buffer_km*1000
  central_buffer<-sf::st_buffer(central_spatial, buffer_dist)
  return(central_buffer)
  }
```

The central point and the kilometers are the arguments needed.

```{r, warning=FALSE}
This_buffer<-create_buffer(central_point=This_point,buffer_km=20)
```

Confirm that the buffer is not a SpatialPolygons.

```{r}
class(This_buffer)
```

# Map

Use ggplot to create the map.\
Use the functions **geom_raster** to plot the SST data.\
The function **geom_polygon** to plot the base maps data.\
The function **geom_polygon** to plot the buffer.\
Adjust to your corresponding area on the **coord_sf**.

```{r, fig.height=4, warning=FALSE}
ggplot() +
 geom_raster(data=SST_dfclean,aes(x=longitude, y=latitude, fill = sst))+ scale_fill_viridis_c(option = "H")+
 geom_polygon(data = world_df, aes(x = long, y = lat, group = group), colour = '#403d39', fill = "#e5e5e5") +
 geom_sf(data = This_buffer, 
          color = "black", 
          fill = NA,  # equivalent to 'transparent'
          linetype = "dashed")+
 geom_point(data=This_point, aes(x=Longitude, y=Latitude), colour = "black", size = 4)+
 coord_sf(xlim = c(-109.7,-109.1),ylim = c(24.6,25.2))
```

# Extract information

Convert the SST data into a SpatialPointsDataFrame.

```{r}
SST_sp <- SST_dfclean
sp::coordinates(SST_sp) <- ~longitude + latitude
sp::proj4string(SST_sp) = sp::CRS("+init=epsg:4326")
```

Use the function **over** from the package **sp** to extract the SST data using the buffer.

```{r}
SST_sp<-sf::st_as_sf(SST_sp)
```

```{r}
#LALI01_1colony_over<- over(LALI01_1colony_points,Colony_buffer)
SST_buffer<-sapply(sf::st_intersects(SST_sp,This_buffer), function(z) if (length(z)==0) NA_integer_ else z[1])
```

The SST values that fall inside the buffer are now added as a column named **sst_inside** in the data frame.

```{r}
SST_dfclean$sst_inside <- as.numeric(SST_buffer)
```

The values that fall inside the buffer are shown as **1**.\
To keep only the data inside the buffer select only those rows with values inside the buffer using the function **filter**.

```{r}
SST_inside<-SST_dfclean %>%
  filter(sst_inside==1)%>%
  drop_na(sst)
```

# Subset

To plot the SST data, select the information from the new data frame **SST_inside** and the function **geom_raster** to plot the information.

```{r, fig.height=4}
ggplot() +
  geom_raster(data=SST_inside,aes(x=longitude, y=latitude, fill = sst))+ scale_fill_viridis_c(option = "H")+
  geom_polygon(data = world_df, aes(x = long, y = lat, group = group), colour = '#403d39', fill = "#e5e5e5") +
  
  geom_sf(data = This_buffer, 
          color = "black", 
          fill = NA,  # equivalent to 'transparent'
          linetype = "dashed")+
  geom_point(data=This_point, aes(x=Longitude, y=Latitude), colour = "black", size = 4)+
  theme(panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill='transparent'))+
  coord_sf(xlim = c(-109.7,-109.1),ylim = c(24.6,25.2))+ 
  NULL
```

Use the values inside the buffer to calculate the average SST

```{r}
mean(SST_inside$sst)
sd(SST_inside$sst)
```

# Export

```{r, echo=FALSE, eval=FALSE}
library(here)
here()
ggsave(ggplot() +
  geom_raster(data=SST_inside,aes(x=longitude, y=latitude, fill = sst))+ scale_fill_viridis_c(option = "H")+
  geom_polygon(data = world_df, aes(x = long, y = lat, group = group), colour = '#403d39', fill = "#e5e5e5") +
  
  geom_sf(data = This_buffer, 
          color = "black", 
          fill = NA,  # equivalent to 'transparent'
          linetype = "dashed")+
  geom_point(data=This_point, aes(x=Longitude, y=Latitude), colour = "black", size = 4)+
  theme(panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill='transparent'))+
  coord_sf(xlim = c(-109.7,-109.1),ylim = c(24.6,25.2))+ 
  NULL,
       filename = paste0(here(),"/Blog/2024/buffer_export.jpg"),
       width = 6, 
       height = 4, 
       units = "in",
       dpi = 300)
```

# ~~~~~~~~~~~~~
# 5. May: Magick

# Intro

This post is to read and edit images in R.

Using R assures that your image would not be compress as happens when you open it with some other programs.

The blog consist of two parts:\
1. Add letter to a picture and export\
2. Merge two pictures into one and export

# Packages

Use the package **magick** to process images in R.

For more information: <https://docs.ropensci.org/magick/articles/intro.html#read-and-write>

To install the package

```{r, eval=FALSE}
install.packages("magick")
```

To call the package

```{r}
library(magick)
```

# Data

Select your directory. In the example, as **My_directory**.

```{r, eval=FALSE}
library(here)
My_directory<-here()
```

```{r, echo=FALSE}
Fig1_link<-('https://github.com/MiriamLL/data_supporting_webpage/blob/main/Blog/2024/MABO.JPG?raw=true')
```

Check if you gave the correct path by using the function **list.files**.\
It should show the images you want to use.

Read images using the function **image_read**.

```{r}
Fig1_original<- image_read(Fig1_link)
```

The function **image_draw** will show the picture in the Viewer.\
Moreover, it gives information on the width and height of the picture.

```{r}
Fig1_draw<-image_draw(Fig1_original)
Fig1_draw
```

# Add text

To add text use the function **image_annotate**

To add the text in the right bottom use the information from the dimensions above to define the xaxis and yaxis.

```{r}
Fig1_text<-image_annotate(Fig1_draw, "(a)",  location = "+1199+999", font = 'Arial', size = 100, color='white')
```

```{r, eval=TRUE, echo=FALSE}
library(here)
My_directory<-here()
```

```{r, eval=TRUE, echo=FALSE}
image_write(Fig1_text, path = paste0(My_directory,"/Blog/2024/Fig1_rb.png"), format = "png")
```

```{r, fig.height=4,echo=FALSE, fig.align='left'}
#list.files(My_directory)
Fig1A_rb<-paste0(My_directory,"/Blog/2024/Fig1_rb.png")
knitr::include_graphics(Fig1A_rb)
```

To add the text in the left top part you can use define the xaxis and yaxis as 100.

```{r}
Fig1_text<-image_annotate(Fig1_draw, "(a)",  location = "+50+50", font = 'Arial', size = 100, color='white')
```

```{r, eval=FALSE, echo=FALSE}
image_write(Fig1_text, path = paste0(My_directory,"/Blog/2024/Fig1_lt.png"), format = "png")
```

```{r, fig.height=4,echo=FALSE, fig.align='left'}
#list.files(My_directory)
Fig1A_lt<-paste0(My_directory,"/Blog/2024/Fig1_lt.png")
knitr::include_graphics(Fig1A_lt)
```

# Export

Use the function **image_write** to export.

```{r, eval=FALSE}
image_write(Fig1_text, path = paste0(My_directory,"/Blog/2024/Fig1_tl.png"), format = "png")
```

# Merge two pictures into one and export

Follow the same steps as above, but with a second picture.

```{r,echo=FALSE}
Fig2_link<-('https://github.com/MiriamLL/data_supporting_webpage/blob/main/Blog/2024/RFBO.JPG?raw=true')
Fig2_original<- image_read(Fig2_link)
```

```{r}
Fig2_draw<-image_draw(Fig2_original)
Fig2_text<-image_annotate(Fig2_draw, "(b)",  location = "+50+50", font = 'Arial', size = 100, color='white')
```

# Add border

Use the function **image_border** to add some space between the pictures.

```{r}
Fig1_border<-image_border(image_background(Fig1_text, "transparent"), "white", "40x10")
Fig1_border
```

```{r}
Fig2_border<-image_border(image_background(Fig2_text, "transparent"), "white", "40x10")
Fig2_border
```

# Combine

Use the function **image_append** to have both pictures side by side.

```{r}
image_append(c(Fig1_border,Fig2_border))
```

# Export

Use the function **image_write** to export the picture.

```{r, eval=TRUE, echo=FALSE}
library(here)
My_directory<-here()
```

```{r, eval=FALSE}
image_write(image_append(c(Fig1_border,Fig2_border)), path = paste0(My_directory,"/Blog/2024/Fig1_append.png"), format = "png")
```

# ~~~~~~~~~~~~~
# 6. June: Distance between points

# Intro

This post is about how to calculate distance from a point

# Data

Identify your reference point

```{r}
reference<-data.frame(name="reference",lat=54.18,long=7.89)
```

Convert it to a spatial object

```{r}
reference_sp<-data.frame(lat=reference$lat,
                         lon=reference$lon)
sp::coordinates(reference_sp)<-~lon+lat
```

Then give the information of the other points to compare

```{r}
colonies_sub<-data.frame(name=c("Ailsa Craig","Barra Head","Bass Rock","Bempton Cliff","Bull Rock","Clare Island", "Fair Isle", "Flannan Isles","Foula","Grassholm Island","Great Saltee", "Helgoland" , "Hermaness", "Ireland’s Eye", "Lambay", "Les Etac/Alderrney","Little Skellig","Mykinesholmur","Noss" ,"Ortac","Rouzic Island","Scar Rocks","St. Kilda","Sula sgeir","Sule Skerry","Sule Stack","Troup Head","Westray"),
colony_lat=c(55.25,56.78,56.08,54.15,51.59,53.80,59.54,58.29,60.13,51.73,52.12,54.18,60.71,53.40,53.49,49.70,51.78,62.10,60.14,49.72,48.90,54.67,57.81,59.10,59.08,59.02,57.69,59.29),
colony_long=c(-5.12,-7.64,-2.64,-0.17,-10.27,-10.00,-1.63,-7.59,-2.07,-5.48,-6.62,7.89,-0.88,-6.06,-6.02,-2.24,-10.51,-7.68,-1.02,-2.29,-3.44,-4.70,-8.56,-6.16,-4.41,-4.51,-2.29,-2.98))
```

```{r}
head(colonies_sub)
```

Convert this data frame into a spatial object

```{r}
colonies_sp<-data.frame(lat=colonies_sub$colony_lat,
                        lon=colonies_sub$colony_long)
sp::coordinates(colonies_sp)<-~lon+lat
```

Give both objects the same coordinate system to make sure they are comparable

```{r}
sp::proj4string(reference_sp)= sp::CRS("+init=epsg:4326")
sp::proj4string(colonies_sp)= sp::CRS("+init=epsg:4326")
```

# Calculate

Using the funcion **distm** from the package geosphere the distance can be calculated. Give the argument distHaversine. Read here what is [Haversine](https://en.wikipedia.org/wiki/Haversine_formula).

```{r}
maxdist_m<-(geosphere::distm(reference_sp,colonies_sp,fun = geosphere::distHaversine))
maxdist_km<-round(maxdist_m/1000,digits=2)
```

Load the package **tidyverse** to make some data wrangling

```{r}
library(tidyverse)
```

Use the function **pivot_longer** to reorder the information

```{r}
distances<-as.data.frame(maxdist_km)%>%
  pivot_longer(cols=c(V1:V28))
```

Assign to the data frame

```{r}
colonies_sub$distance<-distances$value
```

Here are the results from the distance between the point and the reference point

```{r}
head(colonies_sub)
```

# Plot

Load the shapefiles from my package **GermanNorthSea**

```{r}
library(GermanNorthSea)
```

Transform them to the assigned CRS using the package **sf**

```{r}
library(sf)
```

Here I used 4326

```{r}
German_land<-st_transform(GermanNorthSea::German_land, 4326)
German_EEZ<-st_transform(GermanNorthSea::German_EEZ, 4326)
German_SCA<-st_transform(GermanNorthSea::German_natura, 4326)
```

Base plot

To create a plot use **ggplot2**

```{r}
library(ggplot2)
```

First using a base plot of the North Sea, highlighting the economic exclusive zone of Germany

```{r, fig.height= 7}
base_plot<-ggplot()+
  geom_sf(data = German_EEZ, colour = "black", fill= NA, lwd = 0.5,linetype="dashed")+
  geom_sf(data = German_land, colour = '#edf2f4', fill = '#2b2d42')+
  coord_sf(xlim = c(-12, 10),ylim = c(48, 63))+
  scale_x_continuous(breaks = c(-10,-5,0,5,10,20,30),labels = function(x) paste0(x, '\u00B0')) +
  scale_y_continuous(breaks = c(45,50,55,60,65,70,75),labels = function(x) paste0(x, '\u00B0'))  +
  theme_bw()+
  xlab('Longitude')+ylab('Latitude')+
  theme(
  panel.background = element_rect(fill = '#edf2f4'),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),legend.position='none',
  panel.border = element_rect(colour = "black", fill=NA, size=1.5),
  
  legend.spacing.y = unit(0.05, 'cm'),
  legend.text=element_text(size=10),
  legend.background = element_rect(fill='transparent',colour ="transparent"),
  legend.box.background = element_rect(fill='transparent',colour ="transparent"),
  legend.key = element_rect(fill = "transparent", colour = "transparent"),

  axis.text.x = element_text(size=10,vjust = 10,color='#3d5a80'),
  axis.text.y = element_text(color='#3d5a80',size=10,margin = margin(0,-1.00,0,1, unit = 'cm')),
  axis.title = element_blank(),
  axis.ticks.length=unit(-0.20, "cm"),
  )
base_plot
```

# Plot with curves

Add the reference point to be able to plot

```{r}
colonies_sub$reference_long<-reference$lon
colonies_sub$reference_lat<-reference$lat
```

Use the function **geom_curve** to create lines between the reference point and the other points.

```{r, fig.height= 7}
curves_plot<-base_plot+
  geom_curve(data= subset(colonies_sub,colonies_sub$name != 'Helgoland'),
             aes(x = reference_long, y = reference_lat, 
                 xend = colony_long, yend = colony_lat),  
                 color = '#faa307',
                 curvature = 0.05, alpha = 0.5)
curves_plot
```

Add the other points using **geom_point**

```{r, fig.height= 7}
points_plot<-curves_plot+
  geom_point(data=colonies_sub,
             aes(x = colony_long,y= colony_lat), 
             color='#faa307', fill='#ffba08',shape=16,size=2,stroke=1.5)+
  
    geom_point(data=reference,
             aes(x = long,y= lat), 
             color='#ff5400', fill='#ff5400',shape=16,size=4,stroke=1.5)
points_plot
```

```{r, fig.height=7}
letters_plot<-points_plot+
  annotate("text", y = 54.15-0.21, x = -0.17+2.5, 
           label = "Bempton Cliff \n (525 km)",
           size=3.5,color='#0d3b66')+
  annotate("text", y = 56.08-0.21, x = -2.64+2.5, 
           label = "Bass Rock \n (701 km)",
           size=3.5,color='#0d3b66')+
  annotate("text", y = 57.69-0.21, x = -2.29+2.0, 
           label = "Troup Head \n (523 km)",
           size=3.5,color='#0d3b66')+
  annotate("text", y = 54.18-0.21, x = 7.89-1.9, 
           label = "Helgoland \n",
           size=3.5,color='red')+
NULL
letters_plot
```

# Resources

Alternatively using the package [nngeo](https://cran.r-project.org/web/packages/nngeo/index.html)\
Check my package [GermanNorthSea](https://www.miriam-lerma.com/posts/2023-06-02-germannorthsea/)\
How to make maps in R [blogpost](https://www.miriam-lerma.com/posts/2023-03-04-mapping-in-r/)

# Export

```{r, echo=FALSE, eval=FALSE}
library(here)
here()
ggsave(letters_plot,
       filename = paste0(here(),"/distance_export.jpg"),
       width = 6, 
       height = 7, 
       units = "in",
       dpi = 300)
```


# ~~~~~~~~~~~~~
# 7.July: Interpolate

# Intro

This post is about how to linear interpolate points from a trip

# Data

Download my package **sula** to have access to test data and the function **interpolate_trips**.

```{r}
#devtools::install_github("MiriamLL/sula")
library(sula)
```

Inside the package, there is data from a tracked individual from Rapa Nui, Chile

```{r}
ID01_track<-sula::GPS01_trips
```

The data contains Latitude, Longitude, Date, Time, ID and trip_number

```{r}
head(ID01_track)
```

Plot the data using **ggplot** to see whats inside.

```{r}
library(ggplot2)
```

Add the central location for reference.

```{r}
ID01_nest<-data.frame(Longitude=-109.4531,Latitude=-27.20097)
```

In the figure, the original points recorded are presented. Each color represents a different individual, the black dot is the colony.

```{r, fig.width=4, fig.height=4}
Original_track<-ggplot(ID01_track, aes(x=Longitude, y=Latitude, color=trip_number)) + 
  geom_point()+
  theme_bw()+
  ggtitle('Original tracking data')+
  theme_bw()+
  theme(legend.position='none')+
  scale_x_continuous(labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(labels = function(x) paste0(-x, '\u00B0'))  +
  xlab('Longitude')+ylab('Latitude')+
  theme(
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),legend.position='none',
    panel.border = element_rect(colour = "black", fill=NA, size=1.5)
  )+
  geom_point(data=ID01_nest,aes(x = Longitude,y= Latitude), 
             color="black", fill="black",shape=16,size=5,stroke=1.5)
Original_track
```

# Interpolations

For the function, provide the following information:\
1. trip_number, this would separate the trips and interpolate them using a loop. This is important to prevent that locations inside the nest are interpolated.\
2. column_lat, and column_lon, need to be provided with the name of your columns. Here, Latitude and Longitude were the names of my columns. I am providing the Latitude and Longitude in degrees.\
3. column_date, column_time and datetime_format should also be fitting your provided data.\
4. interval, please give the interval in seconds. This interval would depend on the type of data you are providing. Because this is a linear interpolation, be careful to not have large gaps. If you do, I would suggest to remove trips with large gaps before doing the interpolation.

## Interpolate to 1 minute

```{r}
ID01_interpolated1m<-interpolate_trips(GPS_data=ID01_track,
                                     interval='60 sec',
                                     column_date='DateGMT',
                                     column_time='TimeGMT',
                                     column_trip='trip_number',
                                     column_lat='Latitude',
                                     column_lon='Longitude',
                                     datetime_format<-"%d/%m/%Y %H:%M:%S")
```

You can see that after the function you have more data points.\
This is because the original data was collected every 5 minutes, and the interpolation added locations every one minute.

```{r}
nrow(ID01_track)
nrow(ID01_interpolated1m)
```

Plot the interpolated data set to see the difference.

```{r, fig.width=4, fig.height=4}
Interpolated_1m<-ggplot(ID01_interpolated1m, aes(x=Longitude, y=Latitude, color=trip_number)) + 
  geom_point()+
  ggtitle('Interpolated to 1 min')+
  theme_bw()+
  theme(legend.position='none')+
  scale_x_continuous(labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(labels = function(x) paste0(-x, '\u00B0'))  +
  xlab('Longitude')+ylab('Latitude')+
  theme(
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),legend.position='none',
    panel.border = element_rect(colour = "black", fill=NA, size=1.5)
  )+
  geom_point(data=ID01_nest,aes(x = Longitude,y= Latitude), 
             color="black", fill="black",shape=16,size=5,stroke=1.5)
Interpolated_1m
```

## Interpolate to 15 minutes

To interpolate to 15 minutes, change the interval to 900 sec.

```{r}
ID01_interpolated15m<-interpolate_trips(GPS_data=ID01_track,
                                     column_trip='trip_number',
                                     column_lat='Latitude',
                                     column_lon='Longitude',
                                     column_date='DateGMT',
                                     column_time='TimeGMT',
                                     datetime_format<-"%d/%m/%Y %H:%M:%S",
                                     interval='900 sec')
```

You can see that after the function you have less data points.\
This is because the original data was collected every 5 minutes, and the interpolation added locations every 15 minutes.

```{r}
nrow(ID01_track)
nrow(ID01_interpolated15m)
```

Plot to see the difference.

```{r, fig.width=4, fig.height=4}
Interpolated_15m<-ggplot(ID01_interpolated15m, aes(x=Longitude, y=Latitude, color=trip_number)) + 
  geom_point()+
  theme_bw()+
  ggtitle('Interpolated to 15 mins')+
  theme_bw()+
  theme(legend.position='none')+
  scale_x_continuous(labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(labels = function(x) paste0(-x, '\u00B0'))  +
  xlab('Longitude')+ylab('Latitude')+
  theme(
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),legend.position='none',
    panel.border = element_rect(colour = "black", fill=NA, size=1.5)
  )+
  geom_point(data=ID01_nest,aes(x = Longitude,y= Latitude), 
             color="black", fill="black",shape=16,size=5,stroke=1.5)
Interpolated_15m
```

# Compare

Using the package **patchwork** we can see the difference side by side.

```{r}
library(patchwork)
```

```{r, fig.width=10, fig.height=4}
Original_track+Interpolated_1m+Interpolated_15m
```

Select the interpolation interval that fits better your data.

# Further reading

To follow instructions from raw data to interpolation [check this blogpost](https://www.miriam-lerma.com/posts/2022-03-22-interpolation/).\
To access the function click [here](https://github.com/MiriamLL/sula/blob/main/R/interpolate_trips.R). You could copy and adjust the function to your data.

If you want to give it a try, there are other packages to interpolate data:\
- [crawl](https://cran.r-project.org/web/packages/crawl/crawl.pdf)\
- [adehabitatLT](https://cran.r-project.org/web/packages/adehabitatLT/adehabitatLT.pdf)\
- [trip](https://cran.r-project.org/web/packages/trip/trip.pdf)\
- [pastecs](https://cran.r-project.org/web/packages/pastecs/index.html)

# Export

```{r, echo=FALSE, eval=FALSE}
library(here)
here()
ggsave(Original_track+Interpolated_1m+Interpolated_15m,
       filename = paste0(here(),"/Blog/2024/interpolate_export.jpg"),
       width = 10, 
       height = 4, 
       units = "in",
       dpi = 300)
```

# ~~~~~~~~~~~~~
# 8.August: Remove locations


# Intro

When studying animals using GPSs, we often need to remove their central location. Here, I am sharing a function I created to eliminate all the locations within an area.

# Data

For the exercises, test data is from masked boobies. <br> To access the data you have to install the package **sula**: devtools::install_github("MiriamLL/sula")

```{r}
#devtools::install_github("MiriamLL/sula")
library(sula)
```

This data frame contains data from 10 individuals.

```{r}
unique(sula::GPS_raw$IDs)
```

To load it into the environment.

```{r}
GPS_ten<-GPS_raw
```

Plot your data

```{r}
library(tidyverse)
```

Here you can see all the recorded locations using the GPSs.

```{r, fig.width=4, fig.height=4}
Original_plot<-ggplot()+
  geom_point(data = GPS_ten,
            aes(x=Longitude,y = Latitude),color='red', size=0.5)+
  scale_x_continuous(labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(labels = function(x) paste0(-x, '\u00B0'))  +
  xlab('Longitude')+ylab('Latitude')+
  theme(
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),legend.position='none',
    panel.border = element_rect(colour = "black", fill=NA, size=1.5)
  )
Original_plot
```

# Create a buffer

Select a location

```{r}
This_point<-data.frame(Longitude=-109.5,Latitude=-27.2)
```

Create a buffer

I created this function to create a buffer around a point

```{r}
create_buffer<-function(central_point=central_point, buffer_km=buffer_km){
  central_spatial<- sp::SpatialPoints(cbind(central_point$Longitude,central_point$Latitude)) 
  sp::proj4string(central_spatial)= sp::CRS("+init=epsg:4326") 
  central_spatial <- sp::spTransform(central_spatial, sp::CRS("+init=epsg:4326"))
  central_spatial<-sf::st_as_sf(central_spatial)
  buffer_dist<-buffer_km*1000
  central_buffer<-sf::st_buffer(central_spatial, buffer_dist)
  return(central_buffer)
  }
```

The parameters to give are the kilometers and the central point

```{r, warning=FALSE}
This_buffer<-create_buffer(central_point=This_point,buffer_km=20)
```

```{r}
class(This_buffer)
```

Here you can see the buffer you created using the point (or central location)

```{r, fig.width=4, fig.height=4}
Buffer_plot<-ggplot()+
  geom_point(data = GPS_ten,
            aes(x=Longitude,y = Latitude),color='red',size=0.5)+
  geom_point(data=This_point,
             aes(x=Longitude,y=Latitude),color='blue')+
  geom_sf(data=This_buffer,colour='blue', fill='transparent', linetype='dashed')+
  scale_x_continuous(labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(labels = function(x) paste0(-x, '\u00B0'))  +
  xlab('Longitude')+ylab('Latitude')+
  theme(
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),legend.position='none',
    panel.border = element_rect(colour = "black", fill=NA, size=1.5)
  )
Buffer_plot
```

Using the function from the package **sp** you can create an spatial object using your GPS data

```{r}
GPS_sp <- GPS_ten
sp::coordinates(GPS_sp) <- ~Longitude + Latitude
sp::proj4string(GPS_sp) = sp::CRS("+init=epsg:4326")
GPS_sp<-sf::st_as_sf(GPS_sp)
```

```{r}
class(GPS_sp)
class(This_buffer)
```

Here the function over identifies which location intersect with the buffer.

```{r}
GPS_over<-sapply(sf::st_intersects(GPS_sp,This_buffer), function(z) if (length(z)==0) NA_integer_ else z[1])
```

This information can be added as a column in the data frame.

```{r}
GPS_ten$In_or_out <- as.numeric(GPS_over)
```

To remove the locations that were within the buffer you can use the function **filter** and **is.na** from the package **tidyverse**

```{r}
GPS_without <- GPS_ten %>%
  filter(is.na(In_or_out)==TRUE)
```

Here you can see that the locations inside the buffer were removed.

```{r, fig.width=4, fig.height=4}
Filtered_plot<-ggplot()+
  geom_point(data = GPS_without,
            aes(x=Longitude,y = Latitude),color='red',size=0.5)+
  geom_point(data=This_point,
             aes(x=Longitude,y=Latitude),color='blue')+
  geom_sf(data=This_buffer,colour='blue', fill='transparent', linetype='dashed')+
  scale_x_continuous(labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(labels = function(x) paste0(-x, '\u00B0'))  +
  xlab('Longitude')+ylab('Latitude')+
  theme(
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),legend.position='none',
    panel.border = element_rect(colour = "black", fill=NA, size=1.5)
  )
Filtered_plot
```

This function can be used to remove all the nest locations within a distance buffer. You could also replace the polygon (here the buffer) for some land shapefiles.

# Compare

Using the package **patchwork** we can see the difference side by side.

```{r}
library(patchwork)
```

```{r, fig.width=9, fig.height=4}
Original_plot+Buffer_plot+Filtered_plot
```

# Further reading

Other functions that did the job:\
gBuffer from rgeos [deprecated](https://cran.r-project.org/web/packages/rgeos/index.html)\
Read more about [package sf](https://github.com/r-spatial/sf/wiki/migrating)

```{r, echo=FALSE, eval=FALSE}
create_buffer<-function(central_point=central_point, buffer_km=buffer_km){
  central_spatial<- sp::SpatialPoints(cbind(central_point$Longitude,central_point$Latitude)) 
  sp::proj4string(central_spatial)= sp::CRS("+init=epsg:4326") 
  central_spatial <- sp::spTransform(central_spatial, sp::CRS("+init=epsg:4326"))
  buffer_dist<-buffer_km/100
  central_buffer<-rgeos::gBuffer(central_spatial, width=1*buffer_dist)
  return(central_buffer)
}
GPS_ten$In_or_out <- as.numeric(sp::over(GPS_sp,This_buffer))
```

# Export

```{r, echo=FALSE, eval=FALSE}
library(here)
here()
ggsave(Original_plot+Buffer_plot+Filtered_plot,
       filename = paste0(here(),"/Blog/2024/remove_export.jpg"),
       width = 9, 
       height = 4, 
       units = "in",
       dpi = 300)
```

# ~~~~~~~~~~~~~
# 9.September: identify trip

# Intro

This post shows an option to assign a trip number to locations of animals.\
This same method could be use to identify events on spatial data.

# Data

For the exercises, test data is from Masked boobies. <br>

```{r}
#devtools::install_github("MiriamLL/sula")
library(sula)
```

This data frame contains data from 10 individuals.

```{r}
unique(sula::GPS_raw$IDs)
```

Using functions from the package **tidyverse**, one of the individuals can be selected.

```{r}
library(tidyverse)
```

```{r}
GPS_one<-GPS_raw %>% filter(IDs=='GPS08')
```

The original data can be checked using functions from the package **ggplot2**.

```{r}
library(ggplot2)
```

In the plot all locations are presented, but events can not be identify yet.

```{r}
Plot_original<-ggplot(GPS_one, aes(x=Longitude, y=Latitude)) + 
  geom_point()+
  scale_x_continuous(labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(labels = function(x) paste0(-x, '\u00B0'))  +
  ggtitle('Tracking data')+xlab('Longitude')+ylab('Latitude')+
  theme_bw()+
  theme(
    legend.position='none',
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=1.5))
Plot_original
```

# Remove central locations

An event would be defined as each time the animal leaves its central location.

The information about the central location is needed.

```{r}
Central_location<-data.frame(Longitude=-109.4531, Latitude=-27.20097)
```

The locations are to be transform into spatial data.

```{r}
GPS_spatial <- GPS_one
sp::coordinates(GPS_spatial) <- ~Longitude + Latitude
sp::proj4string(GPS_spatial) = sp::CRS("+init=epsg:4326")
GPS_spatial<-sf::st_as_sf(GPS_spatial)
```

The buffer function is explained in the previous [post](https://www.miriam-lerma.com/posts/2024-08-01-removelocs/).\
The function uses the central location to create a buffer.\
A number "1" shows that the location is inside the 1 km buffer and an NA that is outside the buffer.\
The information is added on a column in the data frame.

```{r}
create_buffer<-function(central_point=central_point, buffer_km=buffer_km){
  central_spatial<- sp::SpatialPoints(cbind(central_point$Longitude,central_point$Latitude)) 
  sp::proj4string(central_spatial)= sp::CRS("+init=epsg:4326") 
  central_spatial <- sp::spTransform(central_spatial, sp::CRS("+init=epsg:4326"))
  central_spatial<-sf::st_as_sf(central_spatial)
  buffer_dist<-buffer_km*1000
  central_buffer<-sf::st_buffer(central_spatial, buffer_dist)
  return(central_buffer)
  }
This_buffer<-create_buffer(central_point=Central_location,buffer_km=1)
GPS_over<-sapply(sf::st_intersects(GPS_spatial,This_buffer), function(z) if (length(z)==0) NA_integer_ else z[1])
GPS_one$trip <- as.numeric(GPS_over)
```

The NAs (when the animal is outside the buffer) can be replaced for a 0.\
This is not strictly necessary but facilitates the interpretation of the column.

```{r}
GPS_one$trip[is.na(GPS_one$trip)] <- 0
```

Alternatively, text can be added.

```{r}
GPS_one$trip <- gsub("1", "At_central_locations", GPS_one$trip)
GPS_one$trip <- gsub("0", "At_trip", GPS_one$trip)
```

# Assign trip number

To assign the trip number, the first step was adding a new column with a sequential number.

```{r}
GPS_one$sequential_number<-as.integer(paste(seq(1:as.numeric(nrow(GPS_one)))))
```

Afterwards, all the locations inside the central locations were removed.

```{r}
GPS_trips<-subset(GPS_one, GPS_one$trip == "At_trip")
```

By removing the locations inside the central locations, we end up with gaps in the sequence number.\
Here, I added a number of event everytime that the sequence was broken.

```{r}
GPS_trips$trip_number <- (cumsum(c(1L, diff(GPS_trips$sequential_number)) != 1L))+ 1
```

In case there might be more than ten trips, adding more zeros using the function **str_pad** might be useful.

```{r}
GPS_trips$trip_number <- stringr::str_pad(GPS_trips$trip_number ,  2, pad = "0")
```

For easier interpretation, text can be added.

```{r}
GPS_trips$trip_number<-paste0("Trip_", GPS_trips$trip_number )
```

Using the function **unique** the number of events (or trips) can be checked.

```{r}
unique(GPS_trips$trip_number)
```

# Plot

The different events can now be easily observed by adding the argument **color=trip_number** in the plot.

```{r}
Plot_trips<-ggplot(GPS_trips, aes(x=Longitude, y=Latitude, color=trip_number)) + 
  geom_point()+
  geom_point(data=Central_location,aes(x = Longitude,y= Latitude), 
             color="black", fill="black",shape=16,size=5,stroke=1.5)+
  theme_bw()+
  scale_x_continuous(limits=c(-109.7,-108.9),labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(labels = function(x) paste0(-x, '\u00B0'))  +
  ggtitle('Trip number')+xlab('Longitude')+ylab('Latitude')+labs(color='')+
  theme(
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=1.5),
    legend.background = element_rect(colour = "transparent", fill = "transparent"),
    legend.position = c(0.12,0.75))
Plot_trips
```

# Further reading

Functions **count_trips** from the package sula can be used to run loops.

# Export

```{r, echo=FALSE, eval=FALSE}
library(here)
library(patchwork)
here()
ggsave(Plot_original+Plot_trips,
       filename = paste0(here(),"/Blog/2024/identifytrip_export.jpg"),
       width = 9, 
       height = 4, 
       units = "in",
       dpi = 300)
```

# ~~~~~~~~~~~~~
# 10.October: distance between points

# Intro

Calculate distance between points between consecutive locations.\
One of the uses in animal movements it to calculate how much distance an animal covers on one trip.

# Data

For the exercises, test data is from Masked boobies. <br>

```{r}
#devtools::install_github("MiriamLL/sula")
library(sula)
```

This data frame contains data from one individual.\
It has the column Longitude, Latitude and trip_number.

```{r}
GPS_edited<-sula::GPS_edited
```

Load the package **tidyverse**

```{r}
library(tidyverse)
```

Select one of the trips from this individual.

```{r}
GPS_trip <- GPS_edited %>% filter(trip_number=='trip_4')
```

Using the function **geom_path** you can plot the trajectory.

```{r}
library(ggspatial)
```

Add the annotation layer to see the scale.

```{r, fig.width=4, fig.height=4}
ggplot(GPS_trip, aes(x=Longitude, y=Latitude)) + 
  geom_path()+
  scale_x_continuous(limits=c(-109.6,-109.0),breaks=c(-109.5,-109.4,-109.3,-109.2,-109.1),labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(limits=c(-27.3,-26.7),labels = function(x) paste0(-x, '\u00B0'))  +
  ggtitle('Tracking data')+xlab('Longitude')+ylab('Latitude')+
  theme_bw()+
  theme(
    legend.position='none',
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=1.5))+
    annotation_scale(location = "br",bar_cols = c("#ef233c", "#d90429"),text_col = '#ef233c')
```

# Convert to spatial

Convert to spatial object to assign a reference system.

```{r}
GPS_spatial <- GPS_trip[, c("Longitude", "Latitude")]
GPS_spatial <- sp::SpatialPointsDataFrame(coords = GPS_spatial,data = GPS_trip)
sp::proj4string(GPS_spatial) = sp::CRS("+init=epsg:4326")
```

This function calculates distance between consecutive points.

```{r}
GPS_dist <- sapply(2:nrow(GPS_spatial), function(i) {geosphere::distm(GPS_spatial[i - 1, ], GPS_spatial[i,])})
```

To add it back to the data frame consider that the first point is missing. To solve this add a zero or an NA.

```{r}
GPS_dist<- c(0, GPS_dist)
```

It would show in meters, divide by 1000 to show as kilometers, the function **round** helps to transform to only two decimals.

```{r}
GPS_trip$distances <- round(GPS_dist/1000, 2)
```

```{r, echo=FALSE, eval=FALSE}
write_csv(GPS_trip,"GPS_trip.csv")
```

# Plot

To plot as an spatial object (and get the scale correct) use functions from the package **sf**.

```{r}
library(sf)
```

```{r}
GPS_sf <- st_as_sf(GPS_spatial)
```

Here the scale is showed correctly.

```{r, fig.width=4, fig.height=4}
ggplot(GPS_sf)+
  geom_sf()+
  geom_point(data=GPS_trip, aes(x=Longitude,y=Latitude))+
  geom_text(data=GPS_trip, aes(x=Longitude,y=Latitude,label=distances),hjust=-0.3, vjust=0, size=3)+
  geom_path(data=GPS_trip, aes(x=Longitude,y=Latitude))+
  annotation_scale(location = "br",bar_cols = c("#ef233c", "#d90429"),text_col = '#ef233c')+
  scale_x_continuous(limits=c(-109.6,-109.0),breaks=c(-109.5,-109.4,-109.3,-109.2,-109.1),labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(limits=c(-27.3,-26.7),labels = function(x) paste0(-x, '\u00B0'))  +
  ggtitle('Tracking data')+xlab('Longitude')+ylab('Latitude')+
  theme_bw()+
  theme(
    legend.position='none',
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=1.5))
```

# Plot

The trajectory would be easier to interpreter if we include more elements on the plot.\
To include the begging or the end of the trajectory identify the first and last location.

```{r}
GPS_first<-data.frame(Latitude=first(GPS_trip$Latitude),Longitude=first(GPS_trip$Longitude))
GPS_last<-data.frame(Latitude=last(GPS_trip$Latitude),Longitude=last(GPS_trip$Longitude))
```

Here, the first location is shown as the red triangle.\
The last location is is shown as the green square.

```{r, fig.width=4, fig.height=4}
Plot_distance<-ggplot(GPS_sf)+
  geom_sf()+
  geom_point(data=GPS_trip, aes(x=Longitude,y=Latitude),color='#fb8500')+
  geom_text(data=GPS_trip, aes(x=Longitude,y=Latitude,label=distances),hjust=-0.3, vjust=0, size=3)+
  geom_path(data=GPS_trip, aes(x=Longitude,y=Latitude),color='#ffb703')+
  annotation_scale(location = "br",bar_cols = c("#8ecae6", "#219ebc"),text_col = '#023047')+
  scale_x_continuous(limits=c(-109.6,-109.0),breaks=c(-109.5,-109.4,-109.3,-109.2,-109.1),labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(limits=c(-27.3,-26.7),labels = function(x) paste0(-x, '\u00B0'))  +
  ggtitle('Distance between points')+xlab('Longitude')+ylab('Latitude')+
  theme_bw()+
  theme(
    legend.position='none',
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=1.5))+
  geom_point(data=GPS_first, aes(x=Longitude,y=Latitude),color='#c1121f',fill='#c1121f', shape=17, size=4)+
  geom_point(data=GPS_last, aes(x=Longitude,y=Latitude),color='#4f772d',fill='#4f772d',shape=15,size=4)
Plot_distance
```

# Further reading

The function **dist_points** from the package **sula** runs this function in loop.\
This helps when there are more than one trip and several individuals.

```{r, include=FALSE}
GPS_dist<-dist_puntos(GPS_data = GPS_edited,separador='trip_number')
```

# Export

```{r, echo=FALSE, eval=FALSE}
ggsave(Plot_distance,
       filename = paste0(here(),"/Blog/2024/distancepoints_export.jpg"),
       width = 4, 
       height = 4, 
       units = "in",
       dpi = 300)
```


# ~~~~~~~~~~~~~
# 11.November
# Complete or Incomplete

# Intro

When tracking animals, there may be times when they enter areas with no reception.

If a large part of a trip takes place in such areas, you should consider whether to include this data in your calculations.

For instance, long gaps could lead to an underestimation of the actual distance traveled by the animal.

In this post, I'll walk you through the steps I used to determine whether trips are complete or incomplete.

This post has three parts:\
- Create a gap\
- Calculate gaps\
- Identify if our trips are complete or incomplete

# Data

For the exercises, test data is from masked boobies. <br> To access the data you have to install the package **sula**: devtools::install_github("MiriamLL/sula")

```{r}
library(sula)
```

Use the function select from tidyverse to keep the columns ID, latitude, longitude, date, time, and trip number.

```{r}
library(tidyverse)
Data_1original<-GPS_preparado %>% select(IDs,Latitude,Longitude,DateGMT,TimeGMT,trip_number)
```

# Creating a gap

In this part we would eliminate some locations to create an artificial gap

```{r}
library(sf)
```

Select a location with "low reception" and remove all the locations to create the gaps

```{r}
This_area<-data.frame(Longitude=-109.4,Latitude=-27.6)

create_buffer<-function(central_point=central_point, buffer_km=buffer_km){
  central_spatial<- sp::SpatialPoints(cbind(central_point$Longitude,central_point$Latitude)) 
  sp::proj4string(central_spatial)= sp::CRS("+init=epsg:4326") 
  central_spatial <- sp::spTransform(central_spatial, sp::CRS("+init=epsg:4326"))
  central_spatial<-sf::st_as_sf(central_spatial)
  buffer_dist<-buffer_km*1000
  central_buffer<-sf::st_buffer(central_spatial, buffer_dist)
  return(central_buffer)
}

This_buffer<-create_buffer(central_point=This_area,buffer_km=20)

Data_2spatial <- Data_1original
sp::coordinates(Data_2spatial) <- ~Longitude + Latitude
sp::proj4string(Data_2spatial) = sp::CRS("+init=epsg:4326")
Data_2spatial <-sf::st_as_sf(Data_2spatial)
Data_3creategap<-sapply(sf::st_intersects(Data_2spatial ,This_buffer), function(z) if (length(z)==0) NA_integer_ else z[1])
Data_1original$trip <- as.numeric(Data_3creategap)
Data_4withgap <- Data_1original %>%
  filter(is.na(trip)==TRUE)
```

Plot the area with low reception

```{r, fig.width=4, fig.height=4}
Plot_locations<-ggplot()+
  geom_point(data = Data_4withgap,aes(x=Longitude,y = Latitude),color="#6a994e",size=0.5)+
  scale_x_continuous(labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(labels = function(x) paste0(-x, '\u00B0'))  +
  xlab('Longitude')+ylab('Latitude')+
  theme(
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),legend.position='none',
    panel.border = element_rect(colour = "black", fill=NA, size=1.5))+
  geom_sf(data=This_buffer,colour='blue', fill='transparent', linetype='dashed')+
  geom_text(data=This_area, aes(x=Longitude,y=Latitude,label="No reception"),hjust=+0.46, vjust=0, size=3)
Plot_locations
```

# Calculate gaps

A column in the format POSIXct is needed for this example.

```{r}
Data_4withgap <- Data_4withgap %>%
  mutate(dt= as.POSIXct(strptime(paste0(DateGMT," ",TimeGMT), format="%d/%m/%Y %H:%M:%S")))
```

Checking gaps in time on one individual trip.

```{r}
Data_5oneindividual<-Data_4withgap %>% filter(IDs=='GPS09') %>% filter(trip_number=='trip_2')
```

```{r}
Data_5oneindividual<-Data_5oneindividual %>%
  mutate(times_lag=lag(dt))%>%
  mutate(time_dif=as.numeric(difftime(dt, times_lag, units = "mins")))%>%
  mutate(time_dif_mins = round(time_dif,2))
```

This trip has a large gap.

```{r}
range(Data_5oneindividual$time_dif_mins,na.rm=TRUE)
```

Create a loop to calculate gaps separated by individuals and trips.

```{r}
calculate_gaps<-function (this_data = this_data, 
                          column_datetime = column_datetime, 
                          column_separator = column_separator){
  # to standardize the column names
  this_data$column_separator<- this_data[[column_separator]]
  this_data$column_datetime <- this_data[[column_datetime]]
  trips_list <- split(this_data, this_data$column_separator)
    gaps_list <- list()
    for (i in seq_along(trips_list)) {
        trip_df <- trips_list[[i]]
        trip_df<-trip_df %>%
          mutate(times_lag=lag(dt))%>%
          mutate(time_dif=as.numeric(difftime(dt, times_lag, units = "mins")))%>%
          mutate(time_dif_mins = round(time_dif,2))
        gaps_list[[i]] <- trip_df}
    gaps_df <- do.call("rbind", gaps_list)
    return(gaps_df)
}
```

For this, would help to have an unique ID, which helps separate the individual and their trips.

```{r}
Data_5uniqueIds <- Data_4withgap %>%
  mutate(unique_id=paste0(IDs,'_',trip_number))
```

To run the function, provide the data frame, the information for separating, and the name of the column where the date and time are provided.

```{r}
Data_6uniqueIds<-calculate_gaps(this_data=Data_5uniqueIds,
                                column_separator='unique_id',
                                column_datetime='dt')
```

# Complete or incomplete

By using the function summarise, we can see which trips have large gaps.

```{r}
Data_7gaps<-Data_6uniqueIds %>%
  group_by(unique_id)%>%
  summarise(min_gap=min(time_dif_mins,na.rm = TRUE),
            max_gap=max(time_dif_mins,na.rm = TRUE))
```

Based on the gaps during the trips, the trips could be classified as complete or incomplete.\
For example, here an incomplete trip would be any trip that had a gap of more than 30 minutes.

```{r}
Data_7gaps <- Data_7gaps%>%
  mutate(trip_class = case_when(max_gap <= 30 ~ 'Complete',
                                TRUE ~ 'Incomplete'))
```

Use this table to identify those trips that were incomplete.

```{r}
unique_incomplete<-unique((subset(Data_7gaps,Data_7gaps$trip_class=='Incomplete')$unique_id))
```

Use the information above, to add the classification on the tracking locations.

```{r}
Data_8gaps <- Data_6uniqueIds %>%
  mutate(trip_class = case_when(unique_id %in% unique_incomplete ~ 'Incomplete',
                                TRUE ~ 'Complete'))
```

Plot incomplete trips.

```{r, fig.width=4, fig.height=4}
Plot_trips<-ggplot()+
  geom_point(data = Data_8gaps,aes(x=Longitude,y = Latitude,color=trip_class),size=0.5)+
  scale_x_continuous(labels = function(x) paste0(-x, '\u00B0')) +
  scale_y_continuous(labels = function(x) paste0(-x, '\u00B0'))  +
  xlab('Longitude')+ylab('Latitude')+
  theme(
    panel.background = element_rect(fill = '#edf2f4'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),legend.position='none',
    panel.border = element_rect(colour = "black", fill=NA, size=1.5)
  )+
  geom_sf(data=This_buffer,colour='blue', fill='transparent', linetype='dashed')+
  geom_text(data=This_area, aes(x=Longitude,y=Latitude,label="No reception"),hjust=+0.46, vjust=0, size=3)+
  theme(
    legend.background = element_rect(colour = "transparent", fill = "transparent"),
    legend.position = c(0.22,0.15))+
  scale_colour_manual(name='trip',values = c("#6a994e", "#fca311"))+ 
  guides(color = guide_legend(override.aes = list(size = 2)))
Plot_trips
```

## Export

```{r, echo=FALSE, eval=FALSE}
library(here)
library(patchwork)
Plot_locations+Plot_trips
ggsave(Plot_locations+Plot_trips,
       filename = paste0(here(),"/complete_export.jpg"),
       width = 6, 
       height = 10, 
       units = "in",
       dpi = 300)
```


# ~~~~~~~~~~~~~
# 12.December

```{r, eval=FALSE}
library(sula)
```

```{r}
Data_1original<-GPS_preparado 
```

# Transform

The function **fortify** helps the data frame to be more easily be plotted.

```{r}
library(tidyverse)
```

```{r}
Data_2fortify<-fortify(Data_1original)
```

Make sure your lat and lon are numerical

```{r}
Data_2fortify$lat<-as.numeric(Data_2fortify$Latitude)
Data_2fortify$lon<-as.numeric(Data_2fortify$Longitude)
```

Transform to spatial.

```{r}
library(sf)
```

Select which CRS you will like to use.

```{r}
Data_3spatial <- st_as_sf(Data_2fortify, coords = c("lon", "lat"),crs = 4326, agr = "constant")
```

Select the columns that are of interest.

```{r}
Data_4info<-Data_3spatial[,c("IDs","trip_number","dia_hora")]
```

# Map

Install the package **tmap**.

This package allows you to create interactive maps.

```{r}
library(tmap)
```

For making the map interactive, the mode **view** should be declared.

```{r}
tmap_mode("view") 
```

To create your map with points you can use the arguments **tm_shape** and **tm_dots**

```{r}
tm_shape(Data_4info)+
   tm_dots()
```

To rename the information that shows, you can add the argument **popup.vars**

```{r}
tm_shape(Data_4info)+
   tm_dots(id = "IDs",col="yellow",popup.vars=c("trip_number"="trip_number","dt"="dia_hora"))
```

To separate between individuals you can also include them as separated layers.

```{r}
tm_shape(subset(Data_4info,Data_4info$ID=='GPS01'))+
   tm_dots(id = "IDs",col="yellow",popup.vars=c("trip_number"="trip_number","dt"="dia_hora"))+
  tm_shape(subset(Data_4info,Data_4info$ID=='GPS03'))+
   tm_dots(id = "IDs",col="red",popup.vars=c("trip_number"="trip_number","dt"="dia_hora"))
```

# Interact with you map

```{r, echo=FALSE}
library(fontawesome)
```

Using `r fa("plus", fill = 'black')` you can **zoom-in** to see the points.\
Using `r fa("minus", fill = 'black')` you can **zoom-out** to see the points.

Using `r fa("layer-group", fill = 'black')` you can select which **layers** to show.

Put the cursor on top of the dot `r fa("mouse-pointer", fill = 'black')`, when `r fa("hand-point-up", fill = 'black')` appears click on it and you should be able to see the ID, trip number and dt and time of the point.

```{r}
tm_shape(subset(Data_4info,Data_4info$ID=='GPS01'))+
   tm_dots(id = "IDs",col="yellow",popup.vars=c("trip_number"="trip_number","dt"="dia_hora"))+
  tm_shape(subset(Data_4info,Data_4info$ID=='GPS03'))+
   tm_dots(id = "IDs",col="red",popup.vars=c("trip_number"="trip_number","dt"="dia_hora"))
```

Using tmap might save you the step to export your dataframe, and the import it into another program, particularly if you want to check a particular point and particular information associated to that point.

# Further reading

-   [tmap vignette](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)\
-   [examples of use](https://r-tmap.github.io/tmap/index.html)\
-   [book: Elegant and informative maps with tmap](https://r-tmap.github.io/tmap-book/index.html)\
-   [book: geocomputation with R](https://r.geocompx.org/adv-map)

